---
title: "Assignment 1 - Solution"
author: "Sanja Stanisic, n. 800409"
date: "29 March 2020"
output:
  prettydoc::html_pretty:
    df_print: paged
    highlight: vignette
    theme: architect
    toc: yes
    toc_depth: 5
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
  slidy_presentation:
    highlight: default
  pdf_document:
    toc: yes
    toc_depth: 5
  beamer_presentation:
    colortheme: lily
    fig_caption: no
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    theme: Hannover
    toc: yes
  ioslides_presentation:
    css:
    - css/fonts.css
    - css/custom.css
    - css/title-slide.css
    - css/slide-background.css
    includes:
      before_body: html/title.html
    toc: yes
    transition: default
    widescreen: yes
course: Decision Models
---

```{r setup, include=FALSE}
# Use 'verbatim = TRUE' as chunk option to show chunk code as is
if(require(knitr)==FALSE) install.packages("knitr")
hook_source_def = knit_hooks$get('source')
knit_hooks$set(source = function(x, options){
  if (!is.null(options$verbatim) && options$verbatim){
    opts = gsub(",\\s*verbatim\\s*=\\s*TRUE\\s*", "", options$params.src)
    bef = sprintf('\n\n    ```{r %s}\n', opts, "\n")
    stringr::str_c(bef, paste(knitr:::indent_block(x, "    "), collapse = '\n'), "\n    ```\n")
  } else {
     hook_source_def(x, options)
  }
})
```

## Problem
A trading company is looking for a way to maximize profit per transportation of their goods. The company has a train available with 3 wagons. 

When stocking the wagons they can choose among 4 types of cargo, each with its own specifications. 
How much of each cargo type should be loaded on which wagon in order to maximize profit?

### More Data
|TRAIN WAGON $j$	|WEIGHT CAPACITY (TONNE) $w_j$	|VOLUME CAPACITY ($m^2$) $s_j$|
|-----------------|-------------------------------|-----------------------------|
|(wag) 1 	        |10	                            |5000                         |
|(wag) 2	        |8	                            |4000                         |
|(wag) 3	        |12	                            |8000                         |
  



|CARGO TYPE	$i$ |AVAILABLE (TONNE) $a_i$	|VOLUME ($m^2$) $v_i$	|PROFIT (PER TONNE) $p_i$ |
|---------------|-------------------------|---------------------|-------------------------|
|(cg) 1	        |18	                      |400	                |2000                     |
|(cg) 2	        |10	                      |300	                |2500                     |
|(cg) 3	        |5	                      |200	                |5000                     |
|(cg) 4	        |20	                      |500	                |3500                     |


### Decision Variables

$x_{ij}$ the weight of cargo of type $i$ loaded on a wagon $j$. 

$i \in \{1,2,3,4\}$

$j \in \{1,2,3\}$

Therefore, there are **$12$** decision variables.

For example, **$x_{12}$** stands for the weight (in tonnes) of cargo type **$1$** placed in wagon **$2$**.

### Objective Function

The objective function is to maximize the profit:

**$$max: 2000(x_{11} + x_{12} + x_{13}) \\
          + 2500(x_{21} + x_{22} + x_{23}) \\
          + 5000(x_{31} + x_{32} + x_{33}) \\
          + 3500(x_{41} + x_{42} + x_{43}) \\ $$**

### Constraints

**Weight Capacity Constraints**

Total weight of cargo that can be placed in each wagon: 

1. $\sum_{i=1}^{4} x_{i1} \le 10$  


2. $\sum_{i=1}^{4} x_{i2} \le 8$  

3. $\sum_{i=1}^{4} x_{i3} \le 12$  

**Volume Capacity Constraints**

Total volume of cargo that can be placed in each wagon: 

1. $400 x_{11} + 300 x_{21} + 200 x_{31} + 500 x_{41}  \le 5000$ 

2. $400 x_{12} + 300 x_{22} + 200 x_{32} + 500 x_{42}  \le 4000$  

3. $400 x_{13} + 300 x_{23} + 200 x_{33} + 500 x_{43}  \le 8000$  

**Cargo Avaliability Constraints**

Weight of available cargo of each type: 

1.$x_{11} + x_{12} + x_{13}  \le 18$  

2. $x_{21} + x_{22} + x_{23}  \le 10$   

3. $x_{31} + x_{32} + x_{33}  \le 5$ 

4. $x_{41} + x_{42} + x_{43}  \le 20$   

**Boundaries**

The decision variables should not be negative: 

$x_{ij} \ge 0$,  $\forall i,j$

## Model

```{r, echo=TRUE, eval=TRUE, message=FALSE}
if(require(lpSolveAPI)==FALSE) install.packages("lpSolveAPI")

model = make.lp(0,12)

name.lp(model, "Transportation problem")

```

### Objective Function

The decision variables are mapped to the model according to the following scheme: 

$$ x_{11} \rightarrow 1 \\
   x_{12} \rightarrow 2 \\
   x_{13} \rightarrow 3 \\
   x_{21} \rightarrow 4 \\
   x_{22} \rightarrow 5 \\
   x_{23} \rightarrow 6 \\
   x_{31} \rightarrow 7 \\
   x_{32} \rightarrow 8 \\
   x_{33} \rightarrow 9 \\
   x_{41} \rightarrow 10 \\
   x_{42} \rightarrow 11 \\
   x_{43} \rightarrow 12 $$
   
```{r, echo=TRUE, eval=FALSE, message=FALSE}

lp.control(model, sense = "max")

set.objfn(model, obj = c(2000,2000,2000,2500,2500,2500,5000,5000,5000,3500,3500,3500))
```
### Constraints

```{r, echo=TRUE, eval=TRUE}

add.constraint(model,
               xt = c(1,1,1,1),
               type = "<=",
               rhs = 10,
               indices = c(1,4,7,10))

add.constraint (model,
                xt = c(1,1,1,1),
                type = "<=",
                rhs = 8,
                indices = c(2,5,8,11))

add.constraint (model,
                xt = c(1,1,1,1),
                type = "<=",
                rhs = 12,
                indices = c(3,6,9,12))

add.constraint (model,
                xt = c(400,300,200,500),
                type = "<=",
                rhs = 5000,
                indices = c(1,4,7,10))

add.constraint (model,
                xt = c(400,300,200,500),
                type = "<=",
                rhs = 4000,
                indices = c(2,5,8,11))

add.constraint (model,
                xt = c(400,300,200,500),
                type = "<=",
                rhs = 8000,
                indices = c(3,6,9,12))

add.constraint(model,
               xt = c(1,1,1),
               type = "<=",
               rhs = 18,
               indices = c(1:3))

add.constraint(model,
               xt = c(1,1,1),
               type = "<=",
               rhs = 10,
               indices = c(4:6))

add.constraint(model,
               xt = c(1,1,1),
               type = "<=",
               rhs = 5,
               indices = c(7:9))

add.constraint(model,
               xt = c(1,1,1),
               type = "<=",
               rhs = 20,
               indices = c(10:12))

```

### Boundaries
```{r, echo=TRUE, eval=TRUE}

set.bounds(model,lower = rep(0,12))
```

### Model display
```{r, echo=TRUE, eval=TRUE, message=TRUE}

patch <- function( f, before, after ) { 
  f_text <- capture.output(dput(f))
  g_text <- gsub( before, after, f_text )
  g <- eval( parse( text = g_text ) )
  environment(g) <- environment(f)
  g
}

print.lpExtPtr <- patch( print.lpExtPtr, "8", "16" )

model

```
### Model solving
```{r, echo=TRUE, eval=TRUE}
solve(model)
```

### Results
```{r, echo=TRUE, eval=TRUE}
vars <- get.variables(model)
vars
```
As calculated, the profit is maximized when: 

* $5$ tonnes of cargo type $2$ is placed in wagon n.$1$
* $5$ tonnes of cargo type $3$ is placed in wagon n.$1$
* $8$ tonnes of cargo type $4$ is placed in wagon n.$2$
* $12$ tonnes of cargo type $4$ is paced in wagon n.$3$

```{r, echo=TRUE, eval=TRUE}
obj <- get.objective(model)
obj
```
If the suggested model of cargo displacement is applied, the profit obtained will be \$ $107500$. 

```{r, echo=TRUE, eval=TRUE}
constr <- get.constraints(model)
constr
```

According to this model, in order to maximize the profit, the maximum allowed weight is stocked in all wagons. However, in the first wagon only $2500$ $m^2$ out of available $5000$ $m^2$ is occupied, while in the second wagon the entire volume is occupied, and in the third wagon $6000$ $m^2$ out of $8000$ $m^2$ of avaliable space is occupied. 
The solution suggest that we do not stock the cargo type $1$, we load $5$ out of $10$ available tonnes of cargo type $2$, $5$ out of $5$ available tonnes of cargo type $3$, and $20$ out of $20$ available tonnes of cargo type $4$. 

All of the above, can be summarized as follows: 

```{r, echo=TRUE, eval=TRUE}
primal.sol <-get.primal.solution(model)
primal.sol
```

```{r, echo=TRUE, eval=TRUE}
get.basis(lprec = model, nonbasic = FALSE)  
```

Due to the fact that $10$ slack variables are enlisted at the begininig, the optimal basis consists of the following variables: $C_4, C_{11}, C_{12},S_4, S_5, S_6, S_7, C_5, C_7$ and $S_8$


## Sensitivity analysis

```{r, echo=TRUE, eval=TRUE}
if(require(dplyr)==FALSE) install.packages("dplyr")

# Helper function for sensitivity analysis of objective function
printSensitivityObj2 <- function(model){
  options(scipen=999)
  arg.obj = get.sensitivity.obj(model)
  
  numRows <- length(arg.obj$objfrom)
  symb <- c() 
  for (i in c(1:numRows)) 
    symb[i] <- paste("C", i, sep = "" ) 
   
  obj <- data.frame(Objs = symb, arg.obj)
  a <- mutate(obj,objfrom=replace(objfrom, objfrom < -1.0e4, "-inf"))
  b <- mutate(a,objtill=replace(objtill, objtill > 1.0e4, "inf"))
  result <- ""
  result <- cat("    Objs", "\t", "  Sensitivity", "\n")
  i <- 1
  for (row in 1:nrow(b)) {
    Objs <- b[row, "Objs"]
    objfrom  <- b[row, "objfrom"]
    objtill  <- b[row, "objtill"]
    s <- paste( i, ".\t", Objs, "\t", objfrom, " <= ", Objs, " <= ", objtill)
    result <- cat( result, s, sep = "\n")
    i <- i+1
  }
  print(result)
}
```


```{r, echo=TRUE, eval=TRUE}
printSensitivityObj2(model)
```

The sensitivity analysis of the **Objective Function** suggest that the determined basis will remain the sam if: 

* $-\infty  \ge C_1 \ge 2500$  

* $-\infty  \ge C_2  \ge  2500$ 

* $-\infty  \ge C_3  \ge  2500$   

* $2500 \ge  C_4  \ge  2500$  

* $-\infty  \ge C_5  \ge  2500$   

* $-\infty  \ge C_6  \ge  2500$  

* $5000  \ge  C_7 \ge  \infty$ 

* $-\infty  \ge C_8  \ge  5000$ 

* $-\infty  \ge C_9  \ge  5000$ 

* $-\infty  \ge C_10  \ge  3500$ 

* $3500 \ge C_11  \ge 3500$

* $3500  \ge  C_12  \ge  \infty$ 

```{r, echo=TRUE, eval=TRUE}
get.dual.solution(model)   
```
The result obtained indicates that the array $[2500, 2500, 2500, 0,    0,    0,    0,    0, 2500, 1000]$ represents the shadow prices for the dual variables. 


```{r, echo=TRUE, eval=TRUE}
# Helper function for sensitivity analysis of RHS constraint
printSensitivityRHS2 <- function(model){

  options(scipen=999)
  arg.rhs =get.sensitivity.rhs(model)
  numRows <- length(arg.rhs$duals)
  
  symb <- c() 
  for (i in c(1:numRows)) symb[i] <- paste("B", i, sep = "" ) 
   
  rhs <- data.frame(rhs = symb,arg.rhs)
  
  rhs<-rhs %>%
  mutate(dualsfrom=replace(dualsfrom, dualsfrom < -1.0e4, "-inf")) %>%
  mutate(dualstill=replace(dualstill, dualstill > 1.0e4, "inf")) 
  
  result <- ""
  result <- cat("    RHS", "\t", "Sensitivity", "\n")
  i <- 1
  for (row in 1:nrow(rhs)) {
    vl <- rhs[row, "rhs"]
    dfrom  <- rhs[row, "dualsfrom"]
    dtill  <- rhs[row, "dualstill"]
    duals <-  rhs[row, "duals"]
    s <- paste( i, ".\t", vl, "\t", dfrom, " <= ", vl, " <= ", dtill)
    result <- cat( result, s, sep = "\n")
    i <- i+1
  }
  print(result)
}
```


```{r, echo=TRUE, eval=TRUE}
printSensitivityRHS2(model)
```
The sensitivity analysis of the **RHS** suggests that for the constraints $B_4$ - $B_8$, whose shadow price is zero, a small change  of its value does not affect the solution.

Constraint $B_2$, which represents weight capacity of the wagon n. $2$, is a binding constraint and any change of its value would cause the optimal solution to ne not feasible any more.   

For the remaining constraints, these are the allowed intervals: 

* $5  \le  B_1  \le  15$  
* $12  \le  B_3  \le 16$
* $0  \le  B = 9 \le  10$ 
* $15 \le B_10  \le  20


# Questions about LP

1.	Can an LP model have more than one optimal solution. Is it possible for an LP model to have exactly two optimal solutions? Why or why not?

**Answer:** An LP model can have more than one optimal solution. However, it cannot have exactly two optimal solutions: either we have one or infinite number of solutions. Supposed that A and B are two optimal solutions, then every point of the segment AB have to be an optimal solution.  

2. Are the following objective functions for an LP model equivalent? That is, if they are both used, one at a time, to solve a problem with exactly the same constraints, will the optimal values for $x_1$ and $x_2$ be the same in both cases? Why or why not?

$$
\max 2x_1+3x_2\\
\min -2x_1-3x_2
$$
**Answer:** Yes, they are. They are the same functions with opposite signs, therefore the maximum of the first one is the minimum of the second one.  

3. Which of the following constraints are not linear or cannot be included as a constraint in a linear programming problem? 


a. $2 x_1+x_2-3 x_3 \geq 50$ **Yes**

b. $2 x_1+\sqrt{x_2} \geq 60$ **No**

c. $4 x_1- \frac{1}{2} x_2=75$ **Yes**

d. $\frac{3 x_1+2 x_2x-3 x_3}{x_1+x_2+x_3} \leq 0.9$ **Yes** $x_1+x_2+x_3\gt0$, if variables are non-negative 

e. $3 x_1^2+7 x_2 \leq 45$ **No**

